---
title: "Homework 1"
format: html
editor: visual
---

## Social Network of Jeffrey Epstein

### Loading Data

This project aims to provide a preliminary exploration into the network of actors named in the 'Epstein Files', a collection of documents, images, videos and emails related to the elite U.S. financier and convicted pedophile Jeffrey Epstein. The files detail Epstein's deep influence among his social circle of international elites, and his prolific sex-trafficking of underage girls to members of said circle.

Following the signing of the Epstein Files Transparency Act into law in November of 2025, the U.S. Department of Justice has released over 3 million individual files pertaining to Epstein. Max Andrews created a [visualiser](https://epsteinvisualizer.com) of the social network of entities detailed in the files by extracting structured data from the raw documents using AI (Andrews, 2025). We thank him for his valuable contribution to a case which has been largely abandoned by traditional avenues of justice, left instead to journalists and members of the public to search for the truth and attempt to hold power to account. His data will be used for all following analysis in this project.

```{r}
library(igraph)
library(ggraph)
library(ggplot2)
library(RSQLite)
library(dplyr)
library(kableExtra)

options(timeout = 600)

download.file(
  "https://media.githubusercontent.com/media/maxandrews/Epstein-doc-explorer/main/document_analysis.db",
  "epstein_docs.db",
  mode = "wb")
```

### Plotting Network

```{r}
# Connect
con <- dbConnect(SQLite(), "epstein_docs.db")

# Raw triples
raw <- dbGetQuery(con, "SELECT actor, action, target FROM rdf_triples")
dbDisconnect(con) 

cat("Raw triples:", nrow(raw), "\n")

# Cleaning entity names
alias_map <- c(
  "Jeffrey E." = "Jeffrey Epstein",
  "jeffrey E." = "Jeffrey Epstein",
  "JEE" = "Jeffrey Epstein",
  "Alan M. Dershowitz" = "Alan Dershowitz",
  "Brad Edwards" = "Bradley Edwards",
  "Bradley J. Edwards" = "Bradley Edwards",
  "Bradley James Edwards" = "Bradley Edwards",
  "Edwards" = "Bradley Edwards",
  "Mr. Edwards" = "Bradley Edwards",
  "Virginia Roberts" = "Virginia Giuffre",
  "Virginia Roberts Giuffre" = "Virginia Giuffre",
  "Virginia L. Giuffre" = "Virginia Giuffre",
  "President Clinton" = "Bill Clinton",
  "Donald J. Trump" = "Donald Trump",
  "President Trump" = "Donald Trump",
  "President Obama" = "Barack Obama",
  "Barak" = "Ehud Barak",
  "Ehud" = "Ehud Barak",
  "Stephen K. Bannon" = "Steve Bannon",
  "Lawrence M. Krauss" = "Lawrence Krauss",
  "Dr. Krauss" = "Lawrence Krauss",
  "Lawrence Summers" = "Larry Summers",
  "Alexander Acosta" = "Alex Acosta",
  "R. Alexander Acosta" = "Alex Acosta",
  "Joseph Recarey" = "Detective Recarey",
  "Joe Recarey" = "Detective Recarey",
  "Detective Joseph Recarey" = "Detective Recarey",
  "Larry Visoki" = "Larry Visoski",
  "Lawrance Visoski" = "Larry Visoski",
  "Paul Cassell" = "Paul G. Cassell",
  "Cassell" = "Paul G. Cassell",
  "Mr. Cassell" = "Paul G. Cassell",
  "Kenneth Starr" = "Ken Starr",
  "Kenneth W. Starr" = "Ken Starr",
  "Kathryn Ruemmler" = "Kathy Ruemmler",
  "Jean-Luc Brunel" = "Jean Luc Brunel",
  "Martin G. Weinberg" = "Martin Weinberg",
  "Landon Thomas" = "Landon Thomas Jr.",
  "Landon Thomas, Jr." = "Landon Thomas Jr.",
  "Mr. Scarola" = "Jack Scarola",
  "Federal Bureau of Investigation" = "FBI",
  "Department of Justice" = "Justice Department",
  "U.S. Department of Justice" = "Justice Department",
  "United States Government" = "U.S. Government",
  "United States government" = "U.S. Government",
  "court" = "Court"
)

raw$actor  <- trimws(raw$actor)
raw$target <- trimws(raw$target)

clean <- function(x, map) ifelse(x %in% names(map), map[x], x)
raw$actor  <- clean(raw$actor,  alias_map)
raw$target <- clean(raw$target, alias_map)
raw <- raw[raw$actor != raw$target, ]

cat("After alias cleaning:", nrow(raw), "| Unique actors:", n_distinct(raw$actor), "\n")

# Filtering to top 500 nodes
top_actors <- raw |>
  count(actor, name = "freq") |>
  slice_max(freq, n = 500) |>
  pull(actor)

edges <- raw |>
  filter(actor %in% top_actors, target %in% top_actors) |>
  count(actor, target, name = "weight")

cat("After top-500 filter:", nrow(edges), "edges |", 
    n_distinct(c(edges$actor, edges$target)), "nodes\n")

# Igraph object
colnames(edges)[1:2] <- c("from", "to")
edges <- edges[!is.na(edges$from) & !is.na(edges$to), ]

g <- graph_from_data_frame(edges, directed = FALSE)
g <- simplify(g, remove.multiple = TRUE, remove.loops = TRUE)

cat("iGraph nodes:", vcount(g), "\n")
cat("iGraph edges:", ecount(g), "\n")

# Giant component
components <- decompose(g)
gGC <- components[[which.max(sapply(components, vcount))]]

cat("Giant component nodes:", vcount(gGC), "\n")
cat("Giant component edges:", ecount(gGC), "\n")

# Base plot
plot(gGC,
     vertex.label = NA,
     vertex.size  = 3,
     vertex.color = "steelblue",
     edge.color   = "grey70",
     edge.width   = 0.5)

# ggraph plot
ggraph(gGC, layout = "fr") +
  geom_edge_link(aes(alpha = weight), colour = "grey50") +
  scale_edge_alpha(range = c(0.05, 0.6)) +
  geom_node_point(aes(size = degree(gGC)), colour = "steelblue", alpha = 0.8) +
  scale_size(range = c(1, 8)) +
  geom_node_text(
    aes(label = ifelse(
      degree(gGC) > quantile(degree(gGC), 0.95) | name == "Jeffrey Epstein",
      name, ""
    )),
    size = 2.5, repel = TRUE, max.overlaps = Inf
  ) +
  theme_graph() +
  labs(title = "Epstein Document Network", size = "Degree")
```

## Description of Network

#### Describe the network in 1 or 2 lines. What do the nodes and edges represent? What is the number of nodes and links? What’s the network’s sparsity?

```{r RDF Triples}
#  Top 30 actor/action/target combinations by weight
edges_sample <- raw |>
  count(actor, action, target, name = "weight") |>
  arrange(desc(weight)) |>
  slice_head(n = 30)

print(edges_sample)

raw |>
  count(action, name = "freq") |>
  arrange(desc(freq)) |>
  slice_head(n = 50)
```

```{r network-description}
data.frame(
  Nodes       = vcount(gGC),
  Edges       = ecount(gGC),
  Sparsity    = round(2 * ecount(gGC) / (vcount(gGC) * (vcount(gGC) - 1)), 4),
  Diameter    = diameter(gGC),
  Clustering  = round(transitivity(gGC, type = "global"), 4)
) |> 
  kable(caption = "Epstein Document Network — Summary Statistics") |>
  kable_styling(bootstrap_options = c("stripewd", "hover"))

# Top 5 by degree
sort(degree(gGC), decreasing = TRUE)[1:5]
```

This network captures the actors present within the House Oversight Committee's [recent releases of documents](https://www.justice.gov/epstein) related to highly-connected US financier and child sex offender Jeffrey Epstein, extracted via AI. It is an undirected network constructed from RDF triples of the form actor -\> action -\> target (for example, Jeffrey Epstein -\> sent email to -\> Ghislaine Maxwell) and the edges are weighted by the number of appearances of that specific actor/target pair. The most prevalent action is 'sent an email' followed by 'authored', 'met with' and 'sent message to' which reflects the digital nature of many of the documents.

The network has over 19,000 total entities but, due to project constraints, has been limited to include only the top 500 and focuses exclusively on the giant component. It contains 456 nodes and 2022 edges, with a sparsity of just 0.0195. The five most connected nodes are Jeffrey Epstein, unknown person A, Donald Trump, Alan Dershowitz and unknown person B.

The unknown nodes represent represent entities whose identities have been redacted from the files. Unknown person A represents the first unknown entity extracted from a single document, and unknown person B as the second. These unknown actors includes victims of Epstein, whose identities were removed for privacy and protection, as well as Epstein's co-conspirators, associates, employees and other individuals. The D.O.J. was criticised for its many redaction errors, which exposed names, private details and nude images of some of the victims (CITE), and other instances where the redactions hide the names of people who appear to be other perpetrators. Therefore it is hard to classify the unknown nodes as a singular type of entity.

While analyzing our dataset, we found an issue for our network analysis: the "unknown" labels are not stable across documents. The extraction pipeline treats "unknown person A" as the same entity wherever that string appears, but in reality the label is assigned locally within each document — "unknown person A" in one file is almost certainly a different real person than "unknown person A" in another. As a result, these nodes are a composition of different nodes, collapsing multiple distinct individuals into a single node and artificially inflating their degree and centrality. For this reason, we keep the unknown nodes in the main analysis for transparency, but we included a robustness check in which they are removed, to understand how much they affect our results.

## Degree

#### What is the average degree in the network? And the standard deviation of the degree? Do you think the average degree is a good descriptor in this case?

```{r degree-distribution}
data.frame(
  Mean_Degree = round(mean(degree(gGC)), 2),
  SD_Degree   = round(sd(degree(gGC)), 2),
  Median_Degree = median(degree(gGC))
) |> 
  kable(caption = "Epstein Document Network — Degree") |>
  kable_styling(bootstrap_options = c("striped", "hover"))

quantile(degree(gGC))  # full picture of the distribution
```

The network has an average degree of 8.87, meaning that each actor is connected to around 9 others on average. However, the standard deviation is 19.01, over double that of the mean, indicating a large variation in connectedness amongst actors. A small number of highly connected nodes, such as Jeffrey Epstein's 264 or Donald Trump's 143, drag the mean upwards. A more robust picture of the network's connectedness would be the median of 5.

## Degree distribution

#### Plot the degree distribution in linear-linear scale and in log-log-scale. Which do you think it’s a better suited scale for your network’s degree distribution?

### Degree distribution linear-linear scale

```{r}
ggplot() + geom_histogram(aes(x=degree(gGC,mode="all"))) + labs(x="Degree",y="Density")
```

The linear-linear histogram is strongly right-skewed. This means that the vast majority of nodes have very few connections, while there is a small number of high-degree hubs (as we recognized before, nodes like Jeffrey Epstein or Donald Trump). From this plot we cannot see how the hubs are actually distributed along the tail.

### Degree distribution in log-log scale

```{r}
ggplot() + 
  geom_histogram(aes(x = log10(degree(gGC, mode = "all")))) + 
  scale_y_log10() +
  labs(x = "log10(Degree)", y = "log10(Density)")
```

After applying logarithmic scaling to both degree and density, we can see more clearly the distribution of the hubs. There is a descending pattern in the distribution: a few hubs on the right dominate connectivity, while most nodes have only a few links.

We can conclude that, because of the heterogeneous characteristic of our network, where the maximum degree is over 25x the mean (8.87), the log-log plot is indispensable to properly characterize and analyze the structure.

## Assortativity

### What is the degree assortativity of the network? How do you interpret it?

```{r}
assortativity_degree(gGC)
```

The degree assortativity coefficient is **-0.24**. This means that high-degree nodes tend to connect with low-degree nodes rather than to each other. The network is characterized by the domination of a few highly connected brokers, suggesting that high-degree nodes such as Epstein and Trump act as structural bridges, linking subgroups that would otherwise be disconnected. The network is not structured around a small, tightly united elite of mutually connected hubs, rather, these hubs anchor the network individually, each maintaining ties to distinct clusters of lower-degree nodes. This is a network dominated by central intermediaries who control everything.

## Community detection

#### Using any community detection method, does the network have a community structure? Do the detected communities make sense? Discuss.

First, we try a few methods that are suitable for our network. Because the Epstein network is undirected and weighted, the following methods are selected: `cluster_fast_greedy`, `cluster_louvain` and `cluster_label_prop`

```{r}
set.seed(1111)

fg_epstein <- cluster_fast_greedy(gGC)
louvain_epstein <- cluster_louvain(gGC)
label_prop_epstein <- cluster_label_prop(gGC)

communities_method <- tibble(
  Method        = c("Fast Greedy", "Louvain", "Label Propagation"),
  N_communities = c(length(fg_epstein), length(louvain_epstein), length(label_prop_epstein)),
  Modularity    = c(modularity(fg_epstein), modularity(louvain_epstein), modularity(label_prop_epstein))
) |>
  arrange(desc(Modularity))

communities_method
```

After trying the selected methods, we can interpret that Louvain achieved the highest modularity (0.296) with 13 communities, closely followed by Fast Greedy (0.289) with 15 communities. Both methods detect similar underlying structure, but Louvain performs better in modularity and consolidates the network in few communities. Label Propagation performs poorly, with a modularity of 0.0594, and therefore is not considered a suitable method for our network.

The modularity of 0.296 shows us that the structure of our network is moderate / weak. This is consistent with the earlier findings of the assortativity coefficient, as the dominance of the central hubs softens the boundaries between what could be distinguishable subgroups. In other words, the network has a detectable community structure, but the dominance of hubs prevents communities from being more robustly separated.

```{r}
V(gGC)$community <- as.factor(membership(louvain_epstein))

Louvain_communities <- ggraph(gGC, layout = "fr") +
  geom_edge_link(alpha = 0.15, colour = "grey60") +
  geom_node_point(aes(color = community, size = degree(gGC))) +
  scale_size(range = c(1, 8)) +
  geom_node_text(
    aes(label = ifelse(
      degree(gGC) > quantile(degree(gGC), 0.95) | name == "Jeffrey Epstein",
      name, ""
    )),
    size = 2.5, repel = TRUE, max.overlaps = Inf
  ) +
  theme_graph() +
  labs(title = "Epstein Network — Louvain Communities",
       color = "Community", size = "Degree") +
  theme(legend.position = "right")
Louvain_communities 
```

Even if tehre are communities identified, we can see that these communities are highly concentrated. There is no much distinction in between this communities.

## Centrality

#### Compute the degree and betweenness centrality node scores. Who are the most 5 important nodes in each case? Are they the same? What does this tell you about the structure of the network?

```{r}
cent_degree <- degree(gGC, mode = "all")
cent_degree |>
  sort(decreasing = TRUE) |>
  head(5)
```

```{r}
cent_betw   <- betweenness(gGC, weights = NA)
cent_betw |>
  sort(decreasing = TRUE) |>
  head(5)
```

```{r}
cent_betw_w <- betweenness(gGC, weights = E(gGC)$weight)
cent_betw_w |>
  sort(decreasing = TRUE) |>
  head(5)
```

After computing the degree centrality and betweeness, we can interpret a few things about our network. To begin with, there are some edges in the network that overlap in the three rankings - like Epstein, unknown person A, Trump and Dershowitz. This confirm their central role in the network: they are the most connected nodes and the ones with the largest number of shortest possible paths. But, the rankings diverge in some ways: Epstein ranks first in degree and unweighted betweeness, but second in weighted betweeness. This states that, even if Epstein is structurally the most connected actor, the most intense communication flows are mediated by another intermediary. In contrast, Obama appears only in the unweighted betweennes, suggesting that he connects distinct subgroups through low-frequency interactions.

Additionally, unknown person A has a high degree centrality and both weighted and unweighted betweeness, indicating that a large part of the network is connected to someone redacted.

This pattern reveals that the operation in the Epstein network was based on the fact that high-profile figures are the visible topology and act as structural bridges, but the anonymized actors carry the weight of the recurrent interactions. So, the logic of the Epstein network was to organize small set of central brokers connecting loosely connected subgroups in the "peripheria".

## Testing Clustering Coefficicent

7.  Test whether the clustering coefficient of your network can be statistically explained by the configuration model.

```{r}
# Your observed clustering coefficient
observed_cc <- transitivity(gGC, type = "global")
cat("Observed clustering coefficient:", observed_cc, "\n")

# Generate 1000 random networks using configuration model
# and calculate their clustering coefficients
n_simulations <- 1000
random_cc <- numeric(n_simulations)

for (i in 1:n_simulations) {
  # rewire() randomly rewires edges while preserving degree sequence
  g_random <- rewire(gGC, with = keeping_degseq(niter = ecount(gGC) * 10))
  random_cc[i] <- transitivity(g_random, type = "global")
}

# Plot the null distribution vs your observed value
ggplot(data.frame(cc = random_cc), aes(x = cc)) +
  geom_histogram(bins = 30, fill = "steelblue", colour = "white") +
  geom_vline(xintercept = observed_cc, colour = "red", linewidth = 1.2) +
  labs(
    title = "Clustering Coefficient: Observed vs Configuration Model",
    x = "Clustering Coefficient",
    y = "Count",
    caption = "Red line = observed network"
  ) +
  theme_minimal()

# Calculate p-value — proportion of random networks 
# with clustering >= observed
p_value <- mean(random_cc >= observed_cc)
cat("P-value:", p_value, "\n")
cat("Mean random CC:", mean(random_cc), "\n")
cat("Observed CC:", observed_cc, "\n")
```

The clustering coefficient is statistically higher than what could be expected from a random distribution (p-value = 0), indicating that the network has a real community structure. This reveals that the people in Epstein's network tend to know one another.

## Robustness Check — Filtering Out "Unknown" Nodes

As we said in the begginning of this analysis, nodes labeled as "unknown person", "unknown person A", "unknown person B", etc. are likely a composition of different people. This inflates the apparent degree and centrality of these nodes.

To check how much this distorts our results, we rebuild the network excluding any node whose name matches the pattern "unknown" and recompute the centrality measures and community structure. We decided to keep the entities such as China, FBI, United States, because even if they are not actual physical persons, they are identified in the network.

```{r robustness-filter}
# Identify nodes containing "unknown" (case-insensitive)
unknown_nodes <- V(gGC)$name[grepl("unknown", V(gGC)$name, ignore.case = TRUE)]
cat("Removing", length(unknown_nodes), "nodes matching 'unknown':\n")
print(unknown_nodes)

# Build filtered graph
gGC_filtered <- delete_vertices(gGC, unknown_nodes)

# Keep only the giant component of the filtered graph
gGC_filtered <- decompose(gGC_filtered)[[1]]

# Compare sizes
data.frame(
  Network  = c("Full (with unknowns)", "Filtered (identified only)"),
  Nodes    = c(vcount(gGC), vcount(gGC_filtered)),
  Edges    = c(ecount(gGC), ecount(gGC_filtered)),
  Mean_Deg = c(round(mean(degree(gGC)), 2), round(mean(degree(gGC_filtered)), 2)),
  SD_Deg   = c(round(sd(degree(gGC)), 2),   round(sd(degree(gGC_filtered)), 2))
) |>
  kable(caption = "Network size before and after removing 'unknown' nodes") |>
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r robustness-centrality}
# Top 5 by degree
top_deg <- sort(degree(gGC_filtered), decreasing = TRUE)[1:5]

# Top 5 by betweenness
top_btw <- sort(betweenness(gGC_filtered, normalized = TRUE), decreasing = TRUE)[1:5]

data.frame(
  Rank = 1:5,
  Top_by_Degree      = names(top_deg),
  Degree             = as.integer(top_deg),
  Top_by_Betweenness = names(top_btw),
  Betweenness        = round(top_btw, 4)
) |>
  kable(caption = "Top 5 nodes in the filtered (identified-only) network") |>
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r robustness-communities}
# Rerun Louvain on the filtered network
set.seed(111)
comm_filtered <- cluster_louvain(gGC_filtered)

cat("Number of communities:", length(comm_filtered), "\n")
cat("Modularity:", round(modularity(comm_filtered), 4), "\n")
cat("Community sizes:\n")
print(sizes(comm_filtered))
```

### Discussion

After removing the 23 "unknown" nodes, the filtered network contains 423 nodes and 1,596 edges. Notably, these 23 nodes (5% of the network) accounted for 21% of all edges, confirming that they were have an important impact in our network. The mean degree drops from 8.87 to 7.55 and the standard deviation drops from 19.01 to 15.44, indicating a less skewed degree distribution once this nodes are excluded. The centrality rankings change in informative ways. In the filtered network, Jeffrey Epstein is the clear top hub by degree, followed by Donald Trump and Alan Dershowitz. The betweenness ranking tells a different story: Trump ranks first, above Epstein. This asymmetry suggests that while Epstein is the most connected individual, Trump plays a stronger structural role, linking parts of the network that would otherwise be less connected. Alan Dershowitz appears in the top 3 of both rankings, consistent with his central role in the legal proceedings. "United States" ranks in the top 5 for betweenness but not degree. Louvain detects 15 communities in the filtered network (vs. 13 in the full network) with modularity 0.3065, a moderate value. The fact that removing the composite unknowns produces more communities is revealing: the unknown nodes were acting as bridges that forced distinct subgroups into shared communities. Once removed, these subgroups separate, producing a slightly finer but still moderately modular partition. Overall, the robustness check suggests that the qualitative conclusions from the full network hold, like Epstein, Trump, Dershowitz, and Clinton remain central, but the unknowns were absorbing a share of connectivity and slightly affeecting the community structure.

Nevertheless, we must once again acknowledge a possibility of bias: the "unknown" label often corresponds to victims and less-public individuals whose identities were legally redacted, so by removing them we are also discarding real structural information about the network. This means that neither the full nor the filtered network represents the "full truth": the full network is distorted by the unknowns, while the filtered network is skewed toward public figures. Given the data available, we have aimed to be as meticulous and methodologically transparent as possible.

## References

Andrews, M. (2025). Epstein-doc-explorer: a graph explorer of the Epstein emails \[Dataset\]. In GitHub. https://github.com/maxandrews/Epstein-doc-explorer
